Loop-Aware Optimizations in PyPy's Tracing JIT

### 2011-06-28 talk/iwtc11/paper.tex
comment out XXXs
sprint planning tuesday

\authorinfo{H\aa kan Ardö}
           {Centre for Mathematical Sciences, Lund University}
           {hakan@debian.org}
\authorinfo{Carl Friedrich Bolz}
           {Heinrich-Heine-Universität Düsseldorf}
           {cfbolz@gmx.de}
\authorinfo{Maciej Fijałkowski}
           {}
           {fijall@gmail.com}

@@

Because $i_0$ is loop-invariant, the addition could be moved out of the loop.
However, we want to get this effect using our existing optimization passes
without changing them too much. Simple optimizations with one forward pass
cannot directly get this effect: They just look at the trace without taking
into account that the trace executes many times in a row. Therefore to achieve
loop-invariant code motion, we peel one iteration off the loop before running
the optimizations. This peeling gives the following trace:

@@

This scheme is quite powerful and generalizes to other optimizations than just
common subexpression elimination. It allows simple linear optimization passes to
perform loop-aware optimizations, such as loop-invariant code motion without
changing them at all. All that is needed is to peel off one iteration, then
apply simple one-pass optimizations and make sure that the necessary extra
arguments are inserted into the label of the loop itself and the jumps
afterwards.

This is the key insight of the proposed implementation scheme: Giving an
optimization two iterations together at the same time gives the optimization
enough context to remove operations from the peeled loop, because it detects
that the operation was performed in the preamble already. Thus at runtime these
moved operations are only executed once when entering the loop and the results
are reused in further iterations.

@@

@@

XXX (samuele): the point about the first iteration is hard to understand

@@

In the general case, each allocation-removed object in the jump arguments is exploded into a
vector of variables containing the values of all registered
fields\footnote{This is sometimes called \emph{scalar replacement}.}.
If some of the fields are themselves references to
allocation-removed objects they are recursively exploded
to make the vector contain only concrete variables. Some care has
to be taken to always place the fields in the same order when
performing this explosion. Notation becomes somewhat simpler if also every
concrete variable of the jump arguments is exploded into a vector containing
itself. For
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is concrete} \\
        H^{\left(k\right)} & \text{if $J_k$ is allocation-removed}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all concrete
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
The arguments of the \lstinline{jump} operation of the peeled loop,
$K$, is constructed by inlining $\hat J$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

@@

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than either CPython or Psyco. This is due to the JIT compilation
advantages and optimizations we discussed in previous work
\cite{bolz_allocation_2011, bolz_runtime_2011}. The geometric mean of the
speedup of loop peeling is 70\%, which makes benchmark times
comparable with native-compiled C code. We attribute the performance gap to C code to
the relative immaturity of PyPy's JIT assembler backend as well as missing
optimizations, like instruction scheduling.

@@

The effect of combining a one ass optimization with loop peeling gives
completely standard loop invariant code motion optimizations
\cite{muchnick_advanced_1997}. We do not claim any novelty in the effect, but
think that our implementation scheme is a very simple one.

Mike Pall, the author of LuaJIT\footnote{\texttt{http://luajit.org/}} seems to
have developped the described technique independently. There are no papers about
LuaJIT but the author of it writes on a mailing list: ``The LOOP pass does
synthetic unrolling of the recorded IR, combining copy-substitution with
redundancy elimination to achieve code hoisting. The unrolled and
copy-substituted instructions are simply fed back into the compiler pipeline,
which allows reuse of all optimizations for redundancy elimination. Loop
recurrences are detected on-the-fly and a minimized set of PHIs is generated.''
\cite{pall_luajit_2009}

>>>>

\authorinfo{H\aa kan Ardö}
           {Centre for Mathematical Sciences, Lund University}
           {hakan@debian.org}
\authorinfo{Carl Friedrich Bolz}
           {Heinrich-Heine-Universität Düsseldorf}
           {cfbolz@gmx.de}
\authorinfo{Maciej Fijałkowski}
           {Unaffiliated}
           {fijall@gmail.com}

@@

Because $i_0$ is loop-invariant, the addition could be moved out of the loop.
However, we want to get this effect using our existing optimization passes
without changing them too much. To achieve this, we peel one iteration off the
loop before running the optimizations. This peeling gives the following trace:

@@

This scheme is quite powerful and generalizes to other optimizations than just
common subexpression elimination. It allows simple linear optimization passes to
perform loop-aware optimizations, such as loop-invariant code motion without
changing them at all. All that is needed is to peel off one iteration, then
apply simple one-pass optimizations and make sure that the necessary extra
arguments are inserted into the label of the loop itself and the jumps
afterwards. Giving the optimizations two iterations together
gives the optimization enough context to remove operations from the peeled loop,
because it detects that the operation was performed in the preamble already.

@@

XXX make clear that the preamble is not necessarily the \emph{first} iteration
of a loop

@@

@@

In the general case, each allocation-removed object in the jump arguments is exploded into a
vector of variables containing the values of all registered
fields\footnote{This is sometimes called \emph{scalar replacement}. XXX check
whether that's true}. If some of the fields are themselves references to
allocation-removed objects they are recursively exploded
to make the vector contain only concrete variables. Some care has
to be taken to always place the fields in the same order when
performing this explosion. Notation becomes somewhat simpler if also every
concrete variable of the jump arguments is exploded into a vector containing
itself. For
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is concrete} \\
        H^{\left(k\right)} & \text{if $J_k$ is allocation-removed}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all concrete
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
The arguments of the \lstinline{jump} operation of the peeled loop,
$K$, is constructed by inlining $\hat J$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

@@

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than either CPython or Psyco. This is due to the JIT compilation
advantages and optimizations we discussed in XXX [ref to other paper]. Loop
peeling gives an additional XXX on average, which makes benchmark times
comparable with native-compiled C code. Missing performance we attribute to
the relative immaturity of PyPy's JIT assembler backend as well as missing
optimizations, like instruction scheduling.

@@

All the optimizations presented here are completely standard
\cite{muchnick_advanced_1997}. XXX

Mike Pall, the author of LuaJIT\footnote{\texttt{http://luajit.org/}} seems to
have developped the described technique independently. There are no papers about
LuaJIT but the author of it writes on a mailing list: "The LOOP pass does
synthetic unrolling of the recorded IR, combining copy-substitution with
redundancy elimination to achieve code hoisting. The unrolled and
copy-substituted instructions are simply fed back into the compiler pipeline,
which allows reuse of all optimizations for redundancy elimination. Loop
recurrences are detected on-the-fly and a minimized set of PHIs is generated."
\cite{pall_luajit_2009}
